PLAE enables quick and easy searching across one million ocular transcriptomes

### 2022-03-12 manuscript_00.Rmd
text updates
text tweaks

---
title: 'PLAE enables quick and easy searching across one million ocular transcriptomes'
author:
  - Vinay Swamy:
      institute: bg
  - Zachary Batz:
      institute: nnrl
  - Robert Hufnagel:
      institute: mgog
  - David McGaughey:
      institute:
        - bg
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - bg: Bioinformatics Group, Ophthalmic Genetics & Visual Function Branch, National Eye Institute, National Institutes of Health
  - nnrl: Neurobiology, Neurodegeneration & Repair Laboratory, National Eye Institute, National Institutes of Health
  - mgog: Medical Genetics and Ophthalmic Genomics Unit, National Eye Institute, National Institutes of Health

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: zotero_plaeApp.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To create a high performance reactive web application to query gene expression across cell type, species, study, and other factors. METHODS: We wrote the Shiny web application PLAE (https://plae.nei.nih.gov) and tailored the structure of the underlying data (sc Eye in a Disk, scEiaD) to display scEiaD. RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes., XX cell types, XX ocular tissues, XX body tissues. As a test of the value of this unified pan-eye dataset, we show XX. CONCLUSION: The PLAE v0.80 web app serves the pan-ocular and body dataset, scEiaD. This offers the eye community a powerful and quick means to test hypotheses on gene and transcript expression across 54 body and 19 eye tissues. "
keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
---

@@

## Pan Atlases contain limited retina cell types or lack integration

@@

## Updated the scEiaD backend
Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types. We updated the scEiaD resource for the PLAE v0.80 web app in five important ways. First, we added two chicken retina (*Gallus gallus*) studies. Second we added an ocular outflow tract dataset. Third, we added a human pan-body reference scRNA dataset. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (sup fig XX). Fourth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 300 to 450. Fifth, we carried and harmonnized several common cell type labels from the Tabula Muris dataset to more easily provide non eye cell type comparisons.

After the dataset and quality control updates the scEiaD v02 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (Fig 1).

# Comparison of PLAE to "pan" web apps with retina data
```{r}
feature_table <- cbind(c('Spectacle', 'No' ,'No', ),
                       c('UCSC Cell Browser', 'Yes', 'No', 'https://cells.ucsc.edu' ),
                       c('Tabula Sapiens', 'No', 'N/A', 'https://tabula-sapiens-portal.ds.czbiohub.org'),
                       c('PLAE','Yes','Yes', 'plae.nei.nih.gov'),
                       c('PanglaoDB', 'Yes', 'No'),
                       c('Protein Atlas', 'Yes','No', 'https://www.proteinatlas.org/humanproteome/celltype'))
colnames(feature_table) <- c('Resource Name', 'Independent Datasets', 'Harmonization', 'URL')
```

## PLAE contains over 10 fold more cells and studies than the closest comparable web app

@@

## Rich web functionality

## Differential testing identifies thousands of genes enriched in various ocular cell types

>>>>

---
title: 'PLAE enables quick and easy searching across one million ocular transcriptomes'
author:
  - Vinay Swamy:
      institute: bg
  - Zachary Batz:
      institute: nnrl
  - David McGaughey:
      institute:
        - bg
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - bg: Bioinformatics Group, Ophthalmic Genetics & Visual Function Branch, National Eye Institute, National Institutes of Health
  - nnrl: Neurobiology, Neurodegeneration & Repair Laboratory, National Eye Institute, National Institutes of Health
  - mgog: Medical Genetics and Ophthalmic Genomics Unit, National Eye Institute, National Institutes of Health

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: zotero_plaeApp.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To create a high performance reactive web application to query gene expression across cell type, species, study, and other factors. METHODS: We wrote the web application PLAE (https://plae.nei.nih.gov) and tailored the structure of the underlying data (sc Eye in a Disk, scEiaD) to display scEiaD. RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes., XX cell types, XX ocular tissues, XX body tissues. As a test of the value of this unified pan-eye dataset, we show XX. CONCLUSION: The PLAE v0.80 web app serves the pan-ocular and body dataset, scEiaD. This offers the eye community a powerful and quick means to test hypotheses on gene and transcript expression across 54 body and 19 eye tissues. "
keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
---

@@

## Pan Atlases contain limited ocular data or lack integration

@@

## Updated the scEiaD backend
Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types. We updated the scEiaD resource for the PLAE v0.90 web app in four important ways. First, we added two chicken retina (*Gallus gallus*) studies. Second we added an ocular outflow tract dataset and two more cornea datasets. Third, we added a human pan-body reference scRNA dataset. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (sup fig XX). Fifth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 300 to 450.

After the dataset and quality control updates the scEiaD v02 dataset contains `r meta_filter_$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (Fig 1).

# Comparison of PLAE to "pan" web apps with retina data
```{r}
feature_table <- rbind(c('Spectacle', 'No' ,'No', 'http://singlecell-eye.org'),
                       c('UCSC Cell Browser', 'Yes', 'No', 'https://cells.ucsc.edu' ),
                       c('Tabula Sapiens', 'No', 'N/A', 'https://tabula-sapiens-portal.ds.czbiohub.org'),
                       c('PLAE','Yes','Yes', 'plae.nei.nih.gov'),
                       c('PanglaoDB', 'Yes', 'No', 'https://panglaodb.se'),
                       c('Protein Atlas', 'Yes','No', 'https://www.proteinatlas.org/humanproteome/celltype'))
colnames(feature_table) <- c('Resource Name', 'Independent Datasets', 'Harmonization', 'URL')
```

## PLAE differs from a complementary web app, Spectacle in several important ways

While PLAE and Spectacle both contain large numbers of ocular-related scRNA datasets, they are structured differently. PLAE is built around the scEiaD database, which integrates all datasets together. This allows for queries to be constructed across studies and

@@

## Rich functionality

## Web accessible
